# Tobit Models

The tobit model, also called a censored regression model, is designed to estimate linear relationships between variables when there is either left- or right-censoring in the dependent variable.

- Censoring from above takes place when cases with a value at or above some threshold, all take on the value of that threshold, so that the true value might be equal to the threshold, but it might also be higher. 

- In the case of censoring from below, values those that fall at or below some threshold are censored.

## Censored and Truncated data

In some data sets we do not observe values above or below a certain magnitude, due to a censoring or truncation mechanism. Examples:

- A central bank intervenes to stop an exchange rate falling below or going above certain levels.

- Dividends paid by a company may remain zero until earnings reach some threshold value.

- A government imposes price controls on some goods.

- A survey of only working women, ignoring non-working women

## Install packages

```{r, message=FALSE}
library(ggplot2)
library(GGally)
library(VGAM)
library(stargazer)
```

## Useful data

```{r}
data <- read.csv("data/tobit.csv")
```

The dataset contains 200 observations. 

- The academic aptitude variable is "apt", the reading test scores are "read" and math test scores are "math". 

- The variable prog is the type of program the student is in, it is a categorical (nominal) variable that takes on three values, academic (prog = 1), general (prog = 2), and vocational (prog = 3). 

- The variable id is an identification variable.

Now let’s look at the data descriptively. 

```{r}
summary(data)
```

```{r}
f <- function(x, var, bw = 15) {
  dnorm(x, mean = mean(var), sd(var)) * length(var)  * bw
}
```


```{r}
p <- ggplot(data, aes(x = apt, fill=prog))
```


```{r}
p + stat_bin(binwidth=15) +
  stat_function(fun = f, size = 1,
    args = list(var = data$apt))
```

```{r}
p + stat_bin(binwidth = 1) + stat_function(fun = f, size = 1, args = list(var = data$apt, 
    bw = 1))
```

```{r}
cor(data[, c("read", "math", "apt")])
```

```{r}
ggpairs(data[, c("read", "math", "apt")])
```
```{r}
summary(m <- vglm(apt ~ read + math + prog, tobit(Upper = 800), data = data))
```


